\OurTool{}: Integrative Probing Analysis of Nucleic Acids Empowered by Multiple Accessibility Profiles\\[1em] Supplementary material

### 2019-07-17 IPANEMAP-2019.tex

\maketitle
\begin{abstract}
The manual production of reliable RNA structure models from chemical probing experiments, such as SHAPE probing, benefits from the integration of information derived from multiple protocols and reagents. However, the interpretation of multiple probing profiles remains a complex task, hindering the quality and reproducibility of modeling efforts.

\noindent We introduce \OurTool{}, an automated and robust approach for the computational modeling of RNA structure from multiple probing reactivity profiles. Input profiles can results from experiments based on diverse protocols, reagents, or even over a collection of mutants, and are jointly analyzed to predict the dominant conformations of an RNA.

\noindent \OurTool{} combines statistical sampling, unsupervised machine learning, and multi-objective optimization, to produce secondary structure models that are both stable and well-supported by experimental evidences. The analysis of several reactivity, both publicly available and specifically produced experimentally, demonstrates the good performances of \OurTool, even in a mono probing setting. It confirms the potential of integrating multiple sources of probing data, leading to recommendations for the experimental design of informative probing assays.

@@

\noindent \textbf{Contact:} \href{yann.ponty@lix.polytechnique.fr}{yann.ponty@lix.polytechnique.fr}
%\noindent \textbf{Supplementary information:} Supplementary data available at NAR online.
\end{abstract}

@@

Computationally, the reactivity of a nucleotide is typically used as a proxy for its unpairedness. The last couple of decades have nevertheless seen a series of paradigm shifts in the ways probing information is integrated, somehow mirroring the evolution of {\em ab initio} methods for secondary structure prediction. The seminal work of Mathews~\cite{Mathews2004} used cutoffs to transform reactivity values into hard constraints. Depending on the reagent/enzyme, significantly unreactive or reactive positions were forced to either paired or remain unpaired respectively within predicted structures. However, such \emph{hard} constraints turned out to be very sensitive to the choice of the cutoff, easily leading to artificially unpaired predictions or unsatisfiable sets of constraints. With the advent of SHAPE probing, subsequent methods lifted the requirement of thresholds and integrated reactivities directly as pseudo-energies, supplementing the Turner nearest-neighbor energy model~\cite{Turner2010} within an (pseudo-) energy minimization scheme~\cite{Deigan2009,Zarringhalam2012,Washietl2012}. Such methods can be seen as optimizing a tradeoff between the thermodynamic stability of a model, assessed by the Turner energy model, and its compatibility with the reactivity profile. As a consequence, popular packages for secondary structure prediction, such as \Software{RNAStructure}~\cite{Mathews2004} or the \Software{Vienna} package~\cite{Lorenz2011}, now routinely accept SHAPE-like reactivity profiles as input of their main structure prediction methods. These include the joint folding of aligned RNAs~\cite{Lavender2015}, partition function, statistical sampling and Maximal Expected Accuracy predictions~\cite{Spasic2017}\ldots

@@

Since \OurTool{} supports any number of probing profiles, we consider the Cordero~\emph{et al}~\cite{Cordero2012a} dataset, further described in the previous section. It consists of 6 RNAs of known structures, for which reactivity profiles are available for the CMCT, NMIA and DMS reagents. We executed \OurTool{} with default parameters on each sequence and any subset of the three available conditions. The centroid secondary structure  associated with the largest probability cluster was considered as the final prediction. Predictions were evaluated in term of the geometric mean (GM) metrics, for the sake of comparison with previous studies~\cite{Spasic2017}. As a control experiment, we also report ${\tt RNAfold}$ predictions in presence/absence of probing data, both in energy minimization (MFE) and maximum expected accuracy (MEA) modes. Figure~\ref{cordero1} shows the averaged GM values for all combinations of tools and probing data.

@@

Interestingly, in the absence of probing data, MFE predictions are generally dominant on this dataset, trailed by the MEA, and followed by \OurTool which, in this setting, devolves into the Ding-Lawrence algorithm~\citep{Ding2003}. However, whenever probing reactivities are available, the single centroid returned by \OurTool always achieves higher GM values (Avg: $70\%$) than both MEA (Avg: $62\%$) and MFE  (Avg: $58\%$), whose relative performances depend on the probing reagent. %la moyenne sur les 6 ARNS et sur les 3 reactifs
Interestingly, the quality of MFE and MEA predictions does not systematically benefit from additional probing data. Indeed, for half of the reagents and methods, the average GM obtained in the presence of a single reagent is lower than in the absence of probing data. Also, the impact of single probing data varies greatly across the three reagents, and the GE values of predictions respectively informed by CMCT, DMS and NMIA are ordered increasingly for all approaches, except for a minor reversal of DMS and NMIA in MEA mode.

@@

We executed \OurTool with default parameters, producing 1~000 structures per sample for both software. The Pareto-optimality always implies a single returned cluster/structure in the case of a single probing condition. 
We also executed \Software{RSample} with default parameters, also returning a single structure for each sequence. 
We computed and report in Figure~\ref{fig:Vsrsample} the individual Geometric Means (GM) associated with predicted structures.

@@

\paragraph{Assessment of predictions informed by individual probing conditions (mono).}
%First, we assess the individual performance of each probing condition considered as the sole source of structural constraints.  Then, analyze outperforming experimental conditions that ensure more accurate predictions, and spot \Def{outlier} conditions which, on the contrary, seem to increase the noise level, consequently leading to lower prediction performances for \OurTool{}.

@@

Over the fourteen conditions, we observe a large discrepancy in the capacity of the experimental setup to inform predictions, with MCCs ranging from above 80\% (\NMIA, \NMIAMg, \OneMSevILU and \OneMSevILUMg) to around 60\% (\NAICE, \BzCN, \OneMSevCE and \NMIACE). These values are much better than those achieved in the absence of probing data (51\% MCC), and can be interpreted as indicative of good quality predictions. Indeed, it should be reminded that the MCC is a stringent metrics, taking values between -100\% and 100\%, 0\% being the expected MCC of a "coin tossing" random predictor. The lowest MCC value of 60\% is, in particular, equally consistent with Sensitivity/PPV values of 60\%/60\%, 80\%/46\% or 45\%/82\%.

Unsurprisingly, the presence/absence of Magnesium ions during the probing impacts the predictions, with an observed drop from an average 73\% MCC in the presence of Mg$^\text{2+}$ to 67\% MCC in the absence of Mg$^\text{2+}$. However, while the poorly performing conditions of cluster \Bull{B} are probed in the absence of Mg$^\text{2+}$, some of the most informative profiles (\OneMSevILU and \NMIA) were obtained in absence of Mg$^\text{2+}$.
An even more drastic change of performances can be observed when comparing the MCC of mutation-based protocols (avg 78\%) with stop-based ones (avg 66\%).

@@

\paragraph{Bi-probing mitigates the influence of outliers, but generally fails to improve predictions quality.}
We turn to a systematic exploration of the predictive capacities of bi-probing analyses, based on pairs of probing profiles, and attempt to quantify their impact on \OurTool predictions. For each pair and triplets of conditions, we executed \OurTool using 1000 samples per condition, and considered the first returned structure. We then computed the associated MCC, and compare it with the MCC of the worst performing condition (Min), best-performing condition (Max) and average MCC over the single conditions experiments. A summary of the results over pairs is shown in Figure~\ref{fig:pairwise}.

@@

A first test of this hypothesis consists in executing \OurTool{} on every subset of conditions within clusters, and consider the quality of predictions (MCC). 
Note that our similarity notion and clustering of conditions is purely based on the ensemble features, and does not explicitly consider the MCC of individual conditions. 
As expected, compared to the average MCC over associated mono-probing analyses, a joint consideration of similar conditions only induces limited progress over single probing analyses (between -1\% and +1\%), leading to an modest average improvement (+0.2\% average MCC). This is consistent with the notion that supplementing a probing-based modeling with compatible conditions provides very little additional information, leading to a limited contribution of similar conditions.

@@

We generated 100 uniformly-distributed subsets consisting of 1, 2, 3 and 10 mutants extracted from the Cheng~\emph{et al} dataset~\cite{Cheng2017}, which we supplemented with the original wild-type sequence (WT).
For each set of probing profiles (\Cond{DMS}{mg}{MaP}{ilu}{}), we executed \OurTool{} with default parameters, using a sample size of 1~000 structures. 
We also analyzed the WT sequence alone, reproducing the analysis 100 times to compare the variability across sets of mutants to the one induced by stochastic aspects of our method. 
Finally, we performed a restricted analysis focusing on the WT, supplemented with 2 probing profiles selected from the 20 most similar conditions to the WT , as assessed by the ensemble distance metrics. By comparing our results with those obtained for unrestricted pairs of mutants, we assess the influence of structure variability on the quality of our predictions. We report in Figure~\ref{fig:variantanalysis} the distribution of MCC values associated with the first prediction returned by \OurTool.

@@

Despite its stochastic foundations, \OurTool{} is typically stable in its predictions. We report in Figure~\ref{fig:reproducibility} the MCC values achieved over  10 repetitions of our software for all probing datasets produced in the case of didymium. 
Interestingly, the worst most notable exception to the general observed stability is observed for the probing-free prediction. This confirms the widespread interpretation of SHAPE-induced pseudo-potentials as focusing predictions onto a subset of the Boltzmann ensemble. A similar behavior is observed for the other datasets considered in our study, as seen in Supplementary Material. %In the interest of reproducibility, ipanemap reports, and accepts as an option, the random seed used both within the main software and its non-deterministic dependencies, leading to identical repeats of the experiments.

>>>>

\maketitle
\begin{abstract}
The manual production of reliable RNA structure models from chemical probing experiments, benefits from the integration of information derived from multiple protocols and reagents. However, the interpretation of multiple probing profiles remains a complex task, hindering the quality and reproducibility of modeling efforts.

\noindent We introduce \OurTool{}, an automated method for the computational modeling of RNA structure from multiple probing reactivity profiles. Input profiles can result from experiments based on diverse protocols, reagents, or even a collection of mutants, and are jointly analyzed to predict the dominant conformations of an RNA.

\noindent \OurTool{} combines sampling, unsupervised learning, and multi-optimization, to produce secondary structure models that are both stable and well-supported by experimental evidences. The analysis of several reactivity, both publicly available and specifically produced experimentally, demonstrates the good performances of \OurTool, even in a mono probing setting. It confirms the potential of integrating multiple sources of probing data, leading to recommendations for the experimental design of informative probing assays.

@@

\noindent \textbf{Contact:} \href{yann.ponty@lix.polytechnique.fr}{yann.ponty@lix.polytechnique.fr}

\noindent Supplementary information available at NAR online.
\end{abstract}

@@

Computationally, the reactivity of a nucleotide is typically used as a proxy for the unpairedness of individual nucleotides. The last couple of decades have nevertheless seen a series of paradigm shifts in the ways probing information is integrated, somehow mirroring the evolution of {\em ab initio} methods for secondary structure prediction. The seminal work of Mathews~\cite{Mathews2004} used cutoffs to transform reactivity values into hard constraints. Depending on the reagent/enzyme, significantly unreactive or reactive positions were forced to either paired or remain unpaired respectively within predicted structures. However, such \emph{hard} constraints turned out to be very sensitive to the choice of the cutoff, easily leading to artificially unpaired predictions or unsatisfiable sets of constraints. With the advent of SHAPE probing, subsequent methods lifted the requirement of thresholds and integrated reactivities directly as pseudo-energies, supplementing the Turner nearest-neighbor energy model~\cite{Turner2010} within an (pseudo-) energy minimization scheme~\cite{Deigan2009,Zarringhalam2012,Washietl2012}. Such methods can be seen as optimizing a tradeoff between the thermodynamic stability of a model, assessed by the Turner energy model, and its compatibility with the reactivity profile. As a consequence, popular packages for secondary structure prediction, such as \Software{RNAStructure}~\cite{Mathews2004} or the \Software{Vienna} package~\cite{Lorenz2011}, now routinely accept SHAPE-like reactivity profiles as input of their main structure prediction methods. These include the joint folding of aligned RNAs~\cite{Lavender2015}, partition function, statistical sampling and Maximal Expected Accuracy predictions~\cite{Spasic2017}\ldots

@@

Since \OurTool{} supports any number of probing profiles, we consider the Cordero~\emph{et al}~\cite{Cordero2012a} dataset, further described in the previous section. It consists of 6 RNAs of known structures, for which reactivity profiles are available for the CMCT, NMIA and DMS reagents. We executed \OurTool{} with default parameters on each sequence and any subset of the three available conditions. The centroid secondary structure  associated with the largest probability cluster was considered as the final prediction. Predictions were evaluated in term of the geometric mean (GM) metrics, for the sake of comparison with previous studies~\cite{Spasic2017}. As a control experiment, we also report ${\tt RNAfold}$ predictions in presence/absence of probing data, both in energy minimization (MFE) and maximum expected accuracy (MEA) modes. Figure~\ref{cordero1} shows the averaged GM values for all combinations of tools and probing data (Detailed results can be found in Supp. Tables S1 and S2).

@@

In the absence of probing data, MFE predictions are generally dominant on this dataset, trailed by the MEA, and followed by \OurTool which, in this setting, devolves into the Ding-Lawrence algorithm~\citep{Ding2003}. However, whenever probing reactivities are available, the single centroid returned by \OurTool always achieves higher GM values (Avg: $70\%$) than both MEA (Avg: $62\%$) and MFE  (Avg: $58\%$), whose relative performances depend on the probing reagent. %la moyenne sur les 6 ARNS et sur les 3 reactifs
Interestingly, the quality of MFE and MEA predictions does not systematically benefit from additional probing data. Indeed, for half of the reagents and methods, the average GM obtained in the presence of a single reagent is lower than in the absence of probing data. Also, the impact of single probing data varies greatly across the three reagents, and the GE values of predictions respectively informed by CMCT, DMS and NMIA are ordered increasingly for all approaches, except for a minor reversal of DMS and NMIA in MEA mode.

@@

We executed \OurTool with default parameters, producing 1~000 structures per sample for both software. The Pareto-optimality always implies a single returned cluster/structure in the case of a single probing condition. 
We also executed \Software{RSample} with default parameters, also returning a single structure for each sequence. 
We computed and report in Figure~\ref{fig:Vsrsample} the individual Geometric Means (GM) associated with predicted structures. Sensitivity and PPV reported in Supp. Table S4, while Supp. Table S5 reports the stability of predictions across 10 runs of \OurTool.

@@

\paragraph{Assessment of predictions informed by individual probing conditions (mono probing).}
%First, we assess the individual performance of each probing condition considered as the sole source of structural constraints.  Then, analyze outperforming experimental conditions that ensure more accurate predictions, and spot \Def{outlier} conditions which, on the contrary, seem to increase the noise level, consequently leading to lower prediction performances for \OurTool{}.

@@

Across the sixteen conditions, we observe a large discrepancy in the capacity of the experimental setup to inform predictions, with MCCs ranging from above 80\% (\NMIA, \NMIAMg, \OneMSevILU and \OneMSevILUMg) to around 60\% (\NAICE, \BzCN, \OneMSevCE and \NMIACE). Interestingly the MCC value cannot be trivially anticipated from an assessment of the compatibility of the reactivity scores with the predicted structure, or even with the native structure (see Supp. Table S6).
 These predictive capacities clearly outperform those achieved in the absence of probing data (51\% MCC), and can be interpreted as indicative of good quality predictions. Indeed, remind that the MCC is a stringent metrics, taking values between -100\% and 100\%, 0\% being the expected MCC of a "coin tossing" random predictor. The lowest MCC value of 60\% is, in particular, equally consistent with Sensitivity/PPV values of 60\%/60\%, 80\%/46\% or 45\%/82\%.

Unsurprisingly, the presence/absence of Magnesium ions during the probing impacts the predictions, with an observed drop from an average 73\% MCC in the presence of Mg$^\text{2+}$ to 67\% MCC in the absence of Mg$^\text{2+}$. However, while the poorly performing conditions in Cluster \Bull{B} are probed in the absence of Mg$^\text{2+}$, some of the most informative profiles (\OneMSevILU and \NMIA) were obtained in absence of Mg$^\text{2+}$.
An even more drastic change of performances can be observed when comparing the MCC of mutation-based protocols (avg 78\%) with stop-based ones (avg 66\%).

@@

\paragraph{Bi-probing mitigates the influence of outliers, but generally fails to improve predictions quality.}
We turn to a systematic exploration of the predictive capacities of bi-probing analyses, based on pairs of probing profiles, and attempt to quantify their impact on \OurTool predictions. For each pair and triplets of conditions, we executed \OurTool using 1000 samples per condition, and considered the first returned structure. We then computed the associated MCC, and compare it with the MCC of the worst performing condition (Min), best-performing condition (Max) and average MCC over the single conditions experiments. A summary of the results over pairs is shown in Figure~\ref{fig:pairwise} (see details in Supp. Table S7).

@@

As a test of this hypothesis, we executed \OurTool{} on every subset of conditions within clusters, and report in Figure~\ref{fig:similarconds} the quality (MCC) of predictions for any subset of similar conditions (see Supp. Figure~S1 for associated models). 
Note that our similarity notion and clustering of conditions is purely based on the ensemble features, and does not explicitly consider the MCC of individual conditions. 
As expected, compared to the average MCC over associated mono-probing analyses, a joint consideration of similar conditions only induces limited progress over single probing analyses (between -1\% and +1\%), leading to an modest average improvement (+0.2\% average MCC). This is consistent with the notion that supplementing a probing-based modeling with compatible conditions provides very little additional information, leading to a limited contribution of similar conditions.

@@

We generated 100 uniformly-distributed subsets consisting of 1, 2, 3 and 10 mutants extracted from the Cheng~\emph{et al} dataset~\cite{Cheng2017}, which we supplemented with the original wild-type sequence (WT).
For each set of probing profiles (\Cond{DMS}{mg}{MaP}{ilu}{}), we executed \OurTool{} with default parameters, using a sample size of 1~000 structures. 
We also analyzed the WT sequence alone, reproducing the analysis 100 times to compare the variability across sets of mutants to the one induced by stochastic aspects of our method. 
Finally, we performed a restricted analysis focusing on the WT, supplemented with 2 probing profiles selected from the 20 most similar conditions to the WT , as assessed by the ensemble distance metrics. By comparing our results with those obtained for unrestricted pairs of mutants, we assess the influence of structure variability on the quality of our predictions. We report in Figure~\ref{fig:variantanalysis} the distribution of MCC values associated with the first prediction returned by \OurTool (see Supp. Figure S2 for details).

@@

Despite its stochastic foundations, \OurTool{} is typically stable in its predictions. We report in Figure~\ref{fig:reproducibility} the MCC values achieved over  10 repetitions of our software for all probing datasets produced in the case of didymium. 
Interestingly, the worst most notable exception to the general observed stability is observed for the probing-free prediction. This confirms the widespread interpretation of SHAPE-induced pseudo-potentials as focusing predictions onto a subset of the Boltzmann ensemble. A similar behavior is observed for the other datasets considered in our study, as seen in Supp. Tables S2 and S5. %In the interest of reproducibility, ipanemap reports, and accepts as an option, the random seed used both within the main software and its non-deterministic dependencies, leading to identical repeats of the experiments.

### 2019-07-17 SuppMat.tex

\usepackage{rotating}

@@

\begin{sidewaysfigure}
\newcommand{\MyScale}{.26}

{\centering
	\begin{adjustbox}{max width=\textwidth,max height=\textheight,keepaspectratio}
\begin{tabular}{@{}c|ccccc@{}}

%\toprule
 \includegraphics[scale=\MyScale]{graphs/Supp_structures/native_structure}&
 \includegraphics[scale=\MyScale]{graphs/Supp_structures/1M7ILU_NMIAMg}& \includegraphics[scale=\MyScale]{graphs/Supp_structures/1M7ILUMg_1M7ILU3Mg}&
 \includegraphics[scale=\MyScale]{graphs/Supp_structures/1M7MgCE_NMIAMgCE}
 &
 \includegraphics[scale=\MyScale]{graphs/Supp_structures/1M7CE_NMIACE}
 & \includegraphics[scale=\MyScale]{graphs/Supp_structures/NaiMg_CMCTMg}\\
 \midrule
Native&\OneMSevILU&\OneMSevILUMg&\OneMSevMgCE& \OneMSevCE &\NAIMg\\
structure &+ \NMIAMg&+ \OneMSevILUThreeMg&+ \NMIAMgCE&+ \NMIACE &+ \CMCTMg\\
\end{tabular}
\end{adjustbox}\\}

\caption{Predicted structures with conditions belonging to the same cluster}
\end{sidewaysfigure}

>>>>

\usepackage[figurename=Supplementary~Figure, tablename=Supplementary~Table]{caption}
\DeclareCaptionLabelFormat{cont}{Figure S\arabic{figure}}
\captionsetup[figure]{labelformat=cont}
\DeclareCaptionLabelFormat{contt}{Table S\arabic{table}}
\captionsetup[table]{labelformat=contt}
\captionsetup{labelfont={bf}}

@@

{\newcommand{\MyScale}{.26}

\begin{figure}
{\centering
  \sf	\begin{adjustbox}{max width=\textwidth,max height=\textheight,keepaspectratio}
\begin{tabular}{@{}ccc@{}}

%\toprule
 \includegraphics[scale=\MyScale]{graphs/Supp_structures/native_structure}&
 \includegraphics[scale=\MyScale]{graphs/Supp_structures/1M7ILU_NMIAMg}& 
 \includegraphics[scale=\MyScale]{graphs/Supp_structures/1M7ILUMg_1M7ILU3Mg}\\
Native&\OneMSevILU&\OneMSevILUMg\\ 
structure &+ \NMIAMg&+ \OneMSevILUThreeMg\\[2em]
 \includegraphics[scale=\MyScale]{graphs/Supp_structures/1M7MgCE_NMIAMgCE}
 &
 \includegraphics[scale=\MyScale]{graphs/Supp_structures/1M7CE_NMIACE}
 & \includegraphics[scale=\MyScale]{graphs/Supp_structures/NaiMg_CMCTMg}\\
\OneMSevMgCE& \OneMSevCE &\NAIMg\\
+ \NMIAMgCE&+ \NMIACE &+ \CMCTMg\\
\end{tabular}
\end{adjustbox}\\}

\caption{Predicted structures with conditions belonging to the same cluster}

\end{figure}
